Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife

### 2024-05-09 analysis/paper/paper.Rmd
Writing in progress

institute:
  - ECCC: Environment and Climate Change Canada 
  - FLNR: British Columbia Ministry of Forests 
  - UBC: University of British Columbia 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: Roads.bib
csl: "../templates/canadian-journal-of-forest-research.csl" # Insert path for the bib-style
abstract: |
  Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools suited for this purpose have not been available. Hardy et al. [-@hardy_landis-ii_2023] partially addressed this need with a Landis-II extension. As a complementary and comparable alternative, we provide implementations of simple iterative least cost path and minimum spanning tree methods in an open source R package that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere. Comparison of these methods in a mountainous region of British Columbia clarifies that performance is adequate according to some metrics but not others, that aspatial performance metrics obscure important differences in alorithm behaviour within and outside cutblocks, and that none of these algorithms mimic switchbacks in steep terrain. Method assessment is complicated by errors in available in cutblock and forest road data. Simple iterative least cost path road network simulation methods may be adequate for projecting important impacts on wildlife across large areas, and we hope that more easily useable open source implementations will help integrate resource roads into forest change projections.
keywords: |
  keyword 1; keyword 2; keyword 3
editor_options: 
  chunk_output_type: inline
---

@@

The impacts of roads on wildlife are complex and varied. Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021; @trombulak_review_2000; @fahrig_effects_2009]. In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020; @polfus_identifying_2011; @mumma_predation_2018; @beauchesne_disentangling_2013; @leblond_avoidance_2013; @laurian_interactions_2012;@st-pierre_stairway_2022]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011; @whittington_towns_2022; @proctor_effects_2020]. Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020], and genetics [@habrich_varying_2021]. Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020]. Impacts on wildlife persist after roads are no longer driveable as they continue to be used by wildlife and soil compaction impedes vegetation recovery [@st-pierre_drivers_2021; @lacerte_determinants_2021]. Measures of road impacts are important for informing wildlife management decisions, and there is a need to include roads in ecological forecasting models.

@@

While road projection tools are not typically used in forest landscape models, methods have been developed to support harvest planning [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007], and some forest management planning software suites [e.g. @remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] include road network projection tools. These have occasionally been incorporated into forest landscape models [e.g. @leston_quantifying_2020] but they were not designed for that purpose. The algorithms contained in proprietary closed-source software are difficult to  fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017] used for research because costs of licensing are generally prohibitive, deploying large commercial software suites designed for desktops to cloud or high performance computing resources is difficult or impossible, and closed-source software is often not sufficiently transparent for research. The level of detail required to parameterize forest management planning tools is also often difficult to obtain at landscape scales. Harvest road planning methods and tools [e.g. @anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007; @remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] may produce more accurate road projections but the cost of preparing and running them can be prohibitive in contexts where a rougher approximation of road locations may suffice. There is a need for simple open-source road network projection tools that can be integrated into the open modular platforms that are used for large-scale stochastic landscape simulations [@scheller_design_2007; @barros_empowering_2023; @daniel_state-and-transition_2016].

@@

We acknowledge that using graph analysis methods to simulate roads is not new, and that other variants of iterative least cost path (or source-to-closest-target) and minimum spanning tree algorithms have been proposed and explored [@picard_finding_2006;@anderson_projecting_2004;@stuckelberger_improved_2007;@clark_three-stage_2000]. Anderson and Nelson’s (2004) more complex iterative least cost path method allows for consideration of horizontal and vertical alignment, turn radius and junction angles, and switchback spacing. However, constructing and analyzing graphs for large landscapes is memory intensive and computationally demanding.  We focus here on assessing the adequacy of methods that minimize memory and computational requirements by setting construction costs on a regular isotropic graph from a single input raster. If these simple algorithms can produce adequate outcomes for landscape-level stochastic ecological forecasting then the additional burden of more complex realistic algorithms is not warranted in that context. Both Hardy et al. [-@hardy_forest_2019; -@hardy_landis-ii_2023] and Lochhead et al. [-@lochhead_linking_2022] calculate the cost of constructing a road from one raster pixel to another (represented as graph edge weights) from a single user-supplied cost surface raster that can vary with attributes that alter construction costs such as slope and the presence of water. Here we assess this approach and also propose and assess a simplified method for penalizing steep grades [as suggested by @anderson_projecting_2004] that does not incur the additional computational costs associated with other aspects of their more complex method.

@@

```{r study-area, fig.cap="Road construction cost surface and an elevation surface used to calculate grade penalties. Costs are calculated using a simple model informed by slope and major barriers (Table 1) in the Revelstoke timber supply area in British Colombia, Canada. In both layers, pixels containing existing roads are assigned a value of 0."}

@@

Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for British Columbia. To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the digital road atlas that were within 50 m of forest tenure roads. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990. We assumed everything in the digital road atlas was built before 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021], compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections. We further filtered cutblocks that contain observed roads from those that do not to reduce errors associated with open areas that have been misidentified as cutblocks. Results including all cutblocks are also presented.

Following the Interior Appraisal Manual of BC, we assume the base cost of construction per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table \@ref(tab:cost-tbl)). Existing roads are assigned a construction cost of 0. We compare two different methods for integrating these costs into the cost of connections between adjacent raster cells. In the first method, the cost of the connection between two adjacent cells is simply the mean of a cost raster value at each of those cells, and we refer to it as the “simple cost” method. In the second method, we calculate the cost of a connection between neighbouring cells as a function of the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b). This second method allows us to penalize edge grade as suggested by Anderson and Nelson (2004), and we refer to it as the “grade penalty” method. The road construction cost and elevation rasters consist of 1.7 million pixels with a resolution of 1 ha. We examined sensitivity to our choice of resolution by comparing results from 1 ha and 100 ha cost rasters. We did not consider finer resolutions because stochastic forest landscape models typically simulate dynamics at resolutions >= 1 ha.

```{r cost-tbl}
tbl <- "Cost Type|Value ($ per km)|Source
Slope (%)|16178 + 504*slope|Interior appraisal manual
Major lakes|NA|Major water bodies of BC
Existing roads|0|Digital road atlas and forest tenure roads for BC
"
df <- read.delim(textConnection(tbl),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
names(df) <- unname(as.list(df[1,]))
df <- df[-1,] # remove first row
row.names(df)<-NULL
knitr::kable(df, caption = "A simple model of road construction costs. In the “grade penalty” method, we assign grades >20% a more prohibitive cost of 65000.")
```

@@

The ILCP approaches solve the MTAP by decomposing the problem into several smaller single target access problems (STAPs). For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959]. There are a variety of possible methods for building graphs, choosing branch locations, and ordering targets, and different methods will produce different networks [@anderson_projecting_2004;@hardy_forest_2019; @hardy_landis-ii_2023;@picard_finding_2006]. The iterative least cost path algorithm we use [@lochhead_demo_2018; @lochhead_linking_2022] is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement A). Both the grade penalty and simple cost methods require construction of a simple regular isotropic graph from a single input grid, which is less computationally costly than more complex proposed methods for consideration of curvature and switchbacks [@stuckelberger_improved_2007;@anderson_projecting_2004]. We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest point on the existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing.

Hardy’s method [-@hardy_forest_2019; -@hardy_landis-ii_2023]  is similar to our simple cost ILCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the Euclidean distance from the existing road network. A key difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network. In contrast, our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve least cost path between the two. However, since the cost of construction on existing roads is zero we expect that in practice these two approaches are similar and the algorithm will select the least cost location for the road to connect to the existing road network. It is possible that a sub-optimal road connection location is chosen by our algorithm when the existing road network is not fully connected and the closest point is on an existing road fragment that is not connected to the optimal part of the road network. Within cutblocks, Hardy’s method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.

@@

To help make these road network projection methods more easily and widely useable, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023]. The `roads` package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022], `terra` for manipulating spatial raster data [@hijmans_terra_2023], and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022]. The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].

@@

Maps (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)) help clarify exactly how the projection methods differ from one another, and from real roads. Even in cases where the Hardy method and the grade penalty MST and ILCP methods project similar mean road densities within cutblocks, the Hardy method projects a distinctly different pattern of road tortuosity within cutblocks (Figure \@ref(fig:proj-meth-fig)). The grade penalty method improves realism by making roads follow elevation contours, but all these methods create dead ends, intersections, and abrupt turns that are avoided in the engineering of real roads in steep terrain (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)c). Our projected roads also cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a), but this discrepancy could likely be remedied by better accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms. Finally, it is important to note that some discrepancies between observed and projected road networks may be caused by errors in the observed data, such as cutblocks that do not exist (e.g. \@ref(fig:proj-map-fig)b). We omitted false cutblocks with no associated access roads from our analysis; failure to do this would distort results (Supplement C).

Although some perform better than others, all these methods underproject road density, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)). In other words, these road network projection methods project fewer roads than we observe. Projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project. One reason for underprojection of road density is that all these methods generate straight road segments within pixels. As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure D.1). In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure D.1).

@@

Although the MST method performed better than alternatives according to several metrics (Figure \@ref(fig:aspat-perf-fig)), it also requires more memory and more time to compute on large landscapes (Supplement E). All of these methods require construction of a large graph and therefore have substantial memory requirements; graph size depends on the number of pixels in the cost surface where road construction is possible (i.e. not NA), so options such as decreasing raster resolution or constraining the portion of the landscape where roads are plausible can help reduce memory requirements. Memory requirements and computation time can also be reduced by setting a smaller number of landing targets (e.g. only target the centroid of each cutblock). The Hardy method, run in QGIS, was an order of magnitude faster than any of our methods run in R.

@@

Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools suited for this purpose have not been available. Hardy et al. [-@hardy_forest_2019; -@hardy_landis-ii_2023] partially addressed this need with an extension that can be integrated with other components of the Landis-II modelling framework [e.g. @labadie_global_2023]. Our implementations of iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provide an alternative to Hardy's method that better predicts road patterns in our example landscape, and can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016]. Although projecting roads within ecological forecasting models was beyond the scope of this study, our analysis of performance can assist with method selection in this context.

We evaluated performance of forest road network projection algorithms in a topographically complex region of British Columbia where we anticipated projection would be challenging, and implemented a simplified version of Anderson and Nelson’s (2004) grade penalty method for setting graph edge weights that improves performance in this mountainous region. The simplified grade penalty method does not increase memory or computational requirements, and requires only a few additional lines of code to implement (Supplement A). We see no disadvantages to the approach in regions where slope is an important determinant of road construction cost, and recommend using it. However, we also acknowledge that the important determinants of road construction costs may vary among landscapes, and have also allowed the possibility for users of the `roads` R package to define their own method for determining edge weights from adjacent node values.

@@

Aspatial metrics of overall projection accuracy can be difficult to interpret. To more deeply understand variation in the behaviour of road network projection algorithms we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial patterns. This examination highlighted some errors that could be fixed by better accounting for the cost of water crossing in our weight raster, and other more fundamental limitations and differences among methods. Even in cases where the Hardy and grade penalty methods project similar mean road densities, the projected pattern of roads within cutblocks is quite different, in that the grade penalty roads tend to follow elevation contours, while the Hardy method produces curves (Figure 4). We suspect the latter behaviour is an artifact of searching for nearest destinations \>200 m away on a 100 m grid.

>>>>

institute:
  - ECCC: Environment and Climate Change Canada 
  - UBC: University of British Columbia 
  - FLNR: British Columbia Ministry of Forests 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: Roads.bib
csl: "../templates/canadian-journal-of-forest-research.csl" # Insert path for the bib-style
abstract: |
  Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools suited for this purpose have not been available. Hardy et al. [-@hardy_landis-ii_2023] partially addressed this need with a Landis-II extension. As a complementary and comparable alternative, we provide implementations of simple iterative least cost path and minimum spanning tree methods in an open source R package that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere. Comparison of these methods in a mountainous region of British Columbia clarifies that performance is adequate according to some metrics but not others, that aspatial performance metrics obscure important differences in alorithm behaviour within and outside cutblocks, and that none of these algorithms mimic switchbacks in steep terrain. Method assessment is complicated by errors in available in cutblock and forest road data. Simple iterative least cost path road network simulation methods may be adequate for projecting important impacts on wildlife across large areas, and we hope that more easily useable open source implementations will help integrate resource roads into forest change projections.
keywords: |
  keyword 1; keyword 2; keyword 3
editor_options: 
  chunk_output_type: inline
---

@@

The impacts of roads on wildlife are complex and varied. Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021; @trombulak_review_2000; @fahrig_effects_2009]. In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020; @polfus_identifying_2011; @mumma_predation_2018; @beauchesne_disentangling_2013; @leblond_avoidance_2013; @laurian_interactions_2012;@st-pierre_stairway_2022]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011; @whittington_towns_2022; @proctor_effects_2020]. Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020], and genetics [@habrich_varying_2021]. Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020]. Impacts on wildlife persist after roads are no longer driveable as they continue to be used by wildlife and soil compaction impedes vegetation recovery [@st-pierre_drivers_2021; @lacerte_determinants_2021]. Measures of road impacts are important for informing wildlife management decisions, and there is therefore a need to include roads in ecological forecasting models.

@@

While road projections are not typically integrated with forest landscape models, projection methods have been developed to support harvest planning [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007], and some forest management planning software suites [e.g. @remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] include projection tools. These have occasionally been incorporated into forest landscape models [e.g. @leston_quantifying_2020] but they were not designed for that purpose. The algorithms contained in proprietary closed-source software are difficult to  fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017] used for research because costs of licensing are generally prohibitive, deploying large commercial software suites designed for desktops to cloud or high performance computing resources is difficult or impossible, and closed-source software is often not sufficiently transparent for use in research. The level of detail required to parameterize forest management planning tools is also often difficult to obtain at landscape scales. Harvest road planning methods and tools [e.g. @anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007; @remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] may produce more accurate road projections but the cost of preparing and running them can be prohibitive in contexts where a rougher approximation of road locations may suffice. There is a need for simple open-source road network projection tools that can be integrated into the open modular platforms that are used for large-scale stochastic landscape simulations [@scheller_design_2007; @barros_empowering_2023; @daniel_state-and-transition_2016].

@@

We acknowledge that using graph analysis methods to simulate roads is not new, and that other variants of iterative least cost path (or source-to-closest-target) and minimum spanning tree algorithms have been proposed and explored [@picard_finding_2006;@anderson_projecting_2004;@stuckelberger_improved_2007;@clark_three-stage_2000]. Anderson and Nelson’s (2004) more complex iterative least cost path method allows for consideration of horizontal and vertical alignment, turn radius and junction angles, and switchback spacing. However, constructing and analyzing graphs for large landscapes is memory intensive and computationally demanding.  We focus here on assessing the adequacy of methods that minimize memory and computational requirements by setting construction costs on a regular isotropic graph from a single input raster. If these simple algorithms can produce adequate outcomes for landscape-level stochastic ecological forecasting then the additional burden of more complex realistic algorithms is not warranted in that context. Both Hardy et al. [-@hardy_forest_2019; -@hardy_landis-ii_2023] and Lochhead et al. [-@lochhead_linking_2022] calculate the cost of constructing a road from one raster pixel to another (represented as graph edge weights) from a single user-supplied cost surface raster that can vary with attributes that alter construction costs such as slope and the presence of water. We compare this approach to a simplified method for penalizing steep grades [as suggested by @anderson_projecting_2004] that does not incur the additional computational costs associated with other aspects of their more complex method.

@@

```{r study-area, fig.cap="Road construction cost surface and an elevation surface used to calculate grade penalties. Costs are calculated using a simple model informed by slope and major barriers (Table 1, Supplement A) in the Revelstoke timber supply area in British Colombia, Canada. In both layers, pixels containing existing roads are assigned a value of 0."}

@@

Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for British Columbia. To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the digital road atlas that were within 50 m of forest tenure roads. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990. We assumed everything in the digital road atlas was built before 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021], compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections. We further filtered cutblocks that contain observed roads from those that do not to reduce errors associated with open areas that have been misidentified as cutblocks.

Following the Interior Appraisal Manual of BC, we assume the base cost of construction per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table \@ref(tab:cost-tbl)). Existing roads are assigned a construction cost of 0. We compare two methods for integrating these costs into the cost of connections between adjacent raster cells. In the first "simple cost" method, the cost of the connection between two adjacent cells is simply the mean of a cost raster value at each of those cells. In the second "grade penalty" method (Supplement A), we calculate the cost of a connection between neighbouring cells as a function of the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b), which penalization of edge grades as suggested by Anderson and Nelson (2004). The road construction cost and elevation rasters consist of 1.7 million pixels with a resolution of 1 ha. We examined sensitivity to our choice of resolution by comparing results from 1 ha and 100 ha cost rasters. We did not consider finer resolutions because stochastic forest landscape models typically simulate dynamics at resolutions >= 1 ha.

```{r cost-tbl}
tbl <- "Cost Type|Value ($ per km)|Source
Slope (%)|16178 + 504*slope|Interior appraisal manual
Major lakes|65000|Major water bodies of BC
Existing roads|0|Digital road atlas and forest tenure roads for BC
"
df <- read.delim(textConnection(tbl),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
names(df) <- unname(as.list(df[1,]))
df <- df[-1,] # remove first row
row.names(df)<-NULL
knitr::kable(df, caption = "A simple model of road construction costs. In the “grade penalty” method, we assign grades >20% a more prohibitive cost of 65000, which is also assigned to major water bodies. These barriers are not set to NA because roads must cross barriers to reach a few isolated cutblocks in our example landscape.")
```

@@

The ILCP approaches solve the MTAP by decomposing the problem into several smaller single target access problems (STAPs). For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959]. There are a variety of possible methods for building graphs, choosing branch locations, and ordering targets, and different methods will produce different networks [@anderson_projecting_2004;@hardy_forest_2019; @hardy_landis-ii_2023;@picard_finding_2006]. The ILCP algorithm we use [@lochhead_demo_2018; @lochhead_linking_2022] is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along edges is either the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement A), weighted the distance to neighbouring nodes. Both the grade penalty and simple cost methods require construction of a simple regular isotropic graph from a single input grid, which is less computationally costly than more complex proposed methods for consideration of curvature and switchbacks [@stuckelberger_improved_2007;@anderson_projecting_2004]. We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest point on the existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing.

Hardy’s method [-@hardy_forest_2019; -@hardy_landis-ii_2023]  is similar to our simple cost ILCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the Euclidean distance from the existing road network. A key difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network. In contrast, our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve least cost path between the two. However, since the cost of construction on existing roads is zero we expect that in practice these two approaches are similar and the algorithm will select the least cost location for the road to connect to the existing road network. It is possible that a sub-optimal road connection location is chosen by our algorithm when the existing road network is not fully connected and the closest point is on an existing road fragment that is not connected to the optimal part of the road network. Within cutblocks, Hardy et al's method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.

@@

To help make these road network projection methods more easily and widely useable, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023]. The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022]. The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].

@@

Maps (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)) help clarify exactly how the projection methods differ from one another, and from real roads. Even in cases where the Hardy method and the grade penalty MST and ILCP methods project similar mean road densities within cutblocks, they create distinctly different road patterns within cutblocks, in that the grade penalty method encourages roads to follow slope contours, while the Hardy method creates more curves (Figure \@ref(fig:proj-meth-fig)). Although the grade penalty method improves realism by following elevation contours, all these methods create dead ends, intersections, and abrupt turns that are avoided in the engineering of real roads in steep terrain (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)c). Our projected roads also cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a), but this discrepancy could likely be remedied by better accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms. Finally, it is important to note that some discrepancies between observed and projected road networks may be caused by errors in the observed data, such as cutblocks that do not exist (e.g. \@ref(fig:proj-map-fig)b). We omitted false cutblocks with no associated access roads from our analysis, and note that failure to do this would distort results (Supplement C).

Although some perform better than others, all these methods underproject road density, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)). In other words, these road network projection methods project fewer roads than we observe. Projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project. One reason for underprojection of road density is that all these methods generate straight road segments within pixels. As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure D.1).

@@

Although the MST method performed better than alternatives according to several metrics (Figure \@ref(fig:aspat-perf-fig)), it also requires more memory and time to compute on large landscapes (Supplement E). All of these methods require construction of a large graph with a node for each pixel in the cost surface where road construction is possible (i.e. not NA), and memory requirements increase with the number of nodes in the graph (Supplement E). Compute time also increases with the number of landing targets (Supplement E). The Hardy QGIS plugin was an order of magnitude faster than the MST and ILCP methods implemented in R (Supplement E).

@@

Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools suited for this purpose have not been available. Hardy et al. [-@hardy_forest_2019; -@hardy_landis-ii_2023] partially addressed this need with an extension that can be integrated with other components of the Landis-II modelling framework [e.g. @labadie_global_2023]. Our implementations of iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provide an alternative to Hardy's method that better predicts road patterns in our example landscape, and can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016]. Although projecting roads within ecological forecasting models was beyond the scope of this study, our analysis of performance can assist with method selection in that context.

We evaluated performance of forest road network projection algorithms in a topographically complex region of British Columbia where we anticipated projection would be challenging, and implemented a simplified version of Anderson and Nelson’s (2004) grade penalty method for setting graph edge weights that improves performance in this mountainous region. The simplified grade penalty method does not substantially increase memory or computational requirements, and requires only a few additional lines of code to implement (Supplement A). We see no disadvantages to the approach in regions where slope is an important determinant of road construction cost, and recommend using it. However, we also acknowledge that the important determinants of road construction costs may vary among landscapes, and have allowed the possibility for users of the `roads` R package to specify their own method for determining edge weights from adjacent node values in a weight function.

@@

Aspatial metrics of overall projection accuracy can be difficult to interpret. To more deeply understand variation in the behaviour of road network projection algorithms we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial patterns. This examination highlighted some errors that could be fixed by better accounting for the cost of water crossing in our weight raster, and other more fundamental limitations and differences among methods. Even in cases where the Hardy and grade penalty methods project similar mean road densities, the projected pattern of roads within cutblocks is quite different, in that the grade penalty roads tend to follow elevation contours, while Hardy et al's method produces curves (Figure 4). We suspect the latter behaviour is an artifact of searching for nearest destinations \>200 m away on a 100 m grid.

Although the grade penalty method improves performance by aligning roads with elevation contours, all these methods create dead ends, intersections, and abrupt turns that are avoided in real road networks. We chose not to implement Anderson and Nelson's methods (2004) for penalizing sharp curves, large junction angles and switchback spacing because these complications would increase algorithmic complexity and computational costs. Simple algorithms can recreate some aspects of observed road networks but not others, and more complex realistic algorithms are generally more costly to develop and use. Before investing in additional complexity, it is important to recognize that adequacy depends on the goal. For some purposes, projections from relatively crude cost models and simple algorithms may be adequate. For example, models of caribou movement often include distance to roads as a predictor [e.g. @labadie_global_2023], and a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates among ranges [@johnson_science_2020]. In our example landscape, projections of disturbance footprints (500 m and 1km buffer) and distance to road are reasonably accurate, and for wildlife response models informed by these measures our somewhat crude road projections are likely good enough.
